---
author: Joe Catallini, Eric Bridgeford, Brian Gilbert, Jennifer Xu
output: 
  slidy_presentation
title: Assessing MLB Batter Ability using Statcast Data
header-includes:
  \usepackage{setspace}
  \onehalfspacing 
  \usepackage{amsmath}
  \usepackage{tikz}
  \usepackage{bm}
---

##Introduction

* Baseball has long attracted statisticians as a sport
* Many aspects of baseball can be thought of as quasi-iid trials (pitches, plate appearances, defensive plays, etc.) with specific conditions (count, inning, score, etc.)
* Until recently, statistics used to assess the performance of batters have been outcome based (consider the following 7)
    + Strikeout*
    + Walk/Hit by pitch*
    + Field out
    + Single
    + Double
    + Triple
    + Homerun**
    
*completely determined by pitcher and batter

**almost completely determined by pitcher and batter (stadium effects and inside the park homeruns)

##Brief History of Batter Performance Statistics

* Batting Average (BA): Hits per at-bat
    + Ignores walks and hit by pitches
    + Treats every hit equally
* Slugging Percentage (SLG): Total bases per at-bats
    + Ignores walks and hit by pitches
    + Weights singles, doubles, triples, homeruns (1, 2, 3, 4)
* On-base Percentage (OBP): Total "reach-base-safely" per plate appearance
    + Incorporates walks and hit by pitches
    + Treats every "reach-base-safely" event equally
* OPS
    + OBP + SLG
* Weighted On-base Average (wOBA)
    + Incorporates walks and hit by pitches
    + Regression of historical data assigns linear weights based on expected run values of each event
    + Projects weights to approximate OBP distribution (for interpretability)
    + Weighted average of events per plate appearance
    
##Statcast

* Each of these new statistics may have been an improvement, however they all still rely on outcomes 
* Batted ball outcomes (most notably field outs, singles, doubles, and triples) are influenced by factors outside the batter's control (defensive positioning and proficiency, stadium effects, etc.)
* Beginning in 2015, a new technology called Statcast was introduced to the MLB which gathered data using high speed cameras to track the ball and players
* [Baseball Savant](baseballsavant.mlb.com) has a processed version of the raw data available for public export
* The exportable data includes attributes of each batted ball, the associated pitch, and much, much more
* We have acquired this data from the 2017-2019 season (>500,000 events)

##Project goal

* The primary goal of the project is to utilize Statcast data to assess batter quality from the distribution of their batted balls
* To do this, we will apply methods to predict probabilities of each outcome based on batted ball attributes from the statcast data
    + Random forest
    + Kernel density estimator
    + Bayesian hierarchical models
* Primary predictors of interest
    + Launch speed: velocity of batted ball off the bat
    + Launch angle: angle of batted ball off the bat (relative to plane parallel to the ground)
    + Spray angle: angle of direction of the batted ball (with respect to the field)
    + Sprint speed: top sprint speed of the batter
* Other potential predictors of interest
    + Stadium
    + Defense and defensive alignment: team, infield and outfield shifts
    + Pitch attributes: velocity, spin, location, break, etc.
* We will express batter quality as an expectation of wOBA based on the Statcast data of their batted ball events
* Using this metric, we can evaluate if a given batter's expected wOBA is a better predictor of their future performance than their actual wOBA
    
##DAG

```{r message=F, warning=F, echo=F, fig.align='center'}
library(ggdag)
dag<-dagify(BattedBall~BatterQuality+PitchQuality,
            Outcome~BattedBall+Defense+Stadium+Noise,
            latent = "BatterQuality",
            outcome = "Outcome",
            exposure = c("BattedBall",
                         "PitchQuality",
                         "Stadium",
                         "Defense"))
dag.tidy <- tidy_dagitty(dag)
ggdag_status(dag.tidy, layout="circle", text_col="black") +
  theme_dag() +
  theme(legend.position = "none")
```

##Figure

```{r message=F, warning=F, echo=F, fig.height=9, fig.width=9, fig.align='center'}
library(imager)
img <- load.image('./BattedBalls.png')
plot(img, axes=FALSE)
```
