---
title: "Random Forest Model Analysis and Validation"
author: "Eric Bridgeford"
date: "December 1, 2019"
output: html_document
---

# Model Fitting

## Loading Data

```{r package_load, message=FALSE}
require(rerf)
require(tidyverse)
require(lolR)
require(ggplot2)
require(mltools)
require(parallel)
require(data.table)
```

```{r data_load, eval=FALSE}
batted.dat <- read_csv('../data/cleaned/model_data_batted.csv')
```

```{r data_cc, eval=FALSE}
colnames(batted.dat)
print(sprintf("Fraction of Data Complete: %.4f", mean(complete.cases(batted.dat))))
batted.dat.cc <- batted.dat[complete.cases(batted.dat),]
```

```{r data_clean, eval=FALSE}
# select numeric features
X.num <- batted.dat.cc %>%
  # these are the features which are not 1 of K for K > 2
  select(-outcome, -pitch_name, -stadium, -if_fielding_alignment, -of_fielding_alignment) %>%
  # replace features that have K=2 with numerics 0, 1
  mutate(stand=as.numeric(stand == "R"), p_throws=as.numeric(p_throws == "R"),
         inning_topbot=as.numeric(inning_topbot == "Top"), on_1b_bool=as.numeric(on_1b_bool),
         on_2b_bool=as.numeric(on_2b_bool), on_3b_bool=as.numeric(on_3b_bool)) %>%
  # rename the features according to how we encoded the K=2 features
  rename(batterRighty=stand, pitcherRighty=p_throws, inningTopHalf=inning_topbot, RO1b=on_1b_bool,
         RO2b=on_2b_bool, RO3b=on_3b_bool)

# one-hot encode categorical features where K > 2
X.cat <- batted.dat.cc %>%
  # select the features which are 1 of K for K > 2
  select(pitch_name, stadium, if_fielding_alignment, of_fielding_alignment) %>%
  # turn them into ordered factors
  mutate(pitch_name=factor(pitch_name, levels=unique(pitch_name)),
         stadium=factor(stadium, levels=unique(stadium)),
         if_fielding_alignment=factor(if_fielding_alignment, levels=unique(if_fielding_alignment)),
         of_fielding_alignment=factor(of_fielding_alignment, levels=unique(of_fielding_alignment))) %>%
  # rename with periods so that we can grab the categorical features later since they are OHE'd with _s
  rename(if.fielding.alignment=if_fielding_alignment,
         of.fielding.alignment=of_fielding_alignment) %>%
  # convert to data table so we can OHE
  as.data.table() %>%
  # one-hot encode the entire table
  one_hot()

# paste em together with categoricals all after numerics
X <- cbind(X.num, X.cat)

# identify which categorical features are where
# shift by 1 to account for deleting the "years" column
cat.map <- lapply(c("pitch.name", "stadium", "if.fielding.alignment",
                    "of.fielding.alignment"), function(var) {
  # compute the columns corresponding to a single 1 of K encoding, when K > 2
  as.numeric(which(sapply(colnames(X), function(x) {grepl(var, x)}))) - 1
})

# obtain the outcome variable
Y <- batted.dat.cc %>%
  # factorize
  mutate(outcome=factor(outcome, levels=c("Out", "Single", "Double", "Triple", "Home run"))) %>%
  # extract the column-of-interest
  pull(outcome)
```

Below, we fix a code bug in the prediction code (without this fix, the predictions will be 1:5, not "Single", "Home run", etc.).

```{r correct_bug, eval=FALSE}
fixInNamespace("Predict", ns="rerf")  # bottom content-ful line should be predictions <- labels[predictions]
```

## Prepare Cross-Validation Folds for 50-Fold XV

```{r make_xvsplits, eval=FALSE}
K <- 50  # number of folds for XV
xv.sets <- lol.xval.split(X, Y, k=K)
p <- ncol(X) # number of features in the data
d <- ceiling(sqrt(p)) # number of features to sample at each split
saveRDS(list(X=X, Y=Y, xv.sets=xv.sets, cat.map=cat.map), file='../data/rf/rf_ins.rds')
```

## Cross Validate with 50-fold Cross Validation

Below, we have the batch script that was executed on 10 computers, each featuring between 20 and 40 cores, and between 128 and 256 GB of RAM. Line 113 should be replaced with the suitable index of the machine the script was deployed on.

The results were later aggregated across machines.

```{r execute_rf, eval=FALSE}
require(rerf)
require(tidyverse)
require(lolR)
require(ggplot2)
require(mltools)
require(parallel)
require(data.table)
no_cores <- detectCores() - 1  # number of cores to parallelize over
fixInNamespace("Predict", ns="rerf")
# how many folds should a single machine have?
mult <- 5
# what is the index of the particular machine?
j <- 4
# read in the prepared cross-validation data
inobj <- readRDS('../data/rf/rf_ins.rds')

xv.sets <- inobj$xv.sets[((j-1)*mult + 1):(j*mult)]; X <- inobj$X; Y <- inobj$Y; cat.map <- inobj$cat.map
Y.labs <- unique(Y)
# cross validation loop
res.xv <- lapply(1:length(xv.sets), function(i) {
  print(i)
  # get our cross validation set
  xv.set <- xv.sets[[i]]
  # split X, Y into the training and testing set
  X.train <- X[xv.set$train,] %>% select(-Year); Y.train <- Y[xv.set$train]; 
  X.test <- X[xv.set$test,] %>% select(-Year); Y.test <- Y[xv.set$test]
  # train the forest
  forest <- RerF(X.train, Y.train, num.cores=no_cores, FUN=RandMatRF, cat.map = cat.map)
  # get the predictions on the held-out data
  Y.hat <- rerf::Predict(X.test, forest, num.cores=1L, Xtrain = X.train)
  conf.mtx <- lapply(Y.labs, function(y.true) {
    # obtain indices corresponding to the true label
    Y.test.true <- which(Y.test == y.true)
    n.true <- length(Y.test.true)
    lapply(Y.labs, function(y.pred) {
      Y.test.pred <- which(Y.hat == y.pred)
      # which indices with true label of y.true obtain a prediction of y.pred
      Y.hat.true.pred <- intersect(Y.test.true, Y.test.pred)
      n.true.pred <- length(Y.hat.true.pred)
      
      # record count and percentage
      return(data.frame(fold=i, True=y.true, Prediction=y.pred, n=n.true, n.pred=n.true.pred,
                        percent=n.true.pred/n.true))
    }) %>%
      bind_rows()
  }) %>%
    bind_rows()
  acc.rt <- mean(Y.hat == Y.test)
  # compare accuracy against the chance classifier which just guesses the maximally present class
  # in the training data
  labs.ct <- (Y.train %>% table())
  Y.max <- names(labs.ct)[which(labs.ct == max(labs.ct))]
  dat.overall <- data.frame(fold=i, accuracy=acc.rt, chance=mean(Y.test == Y.max))
  return(list(conf.mtx=conf.mtx, dat.overall=dat.overall))
})
# aggregate the confusion matrix results
conf.mtx <- lapply(res.xv, function(res) res$conf.mtx) %>%
  bind_rows()
# aggregate the overall results
dat.overall <- lapply(res.xv, function(res) res$dat.overall) %>%
  bind_rows()
# save
saveRDS(list(conf.mtx=conf.mtx, overall=dat.overall),
        sprintf('../data/rf/rf_results_%d.rds', j))
```

## Compute Feature Importances on Full Dataset

```{r feat_imp, eval=FALSE}
forest <- RerF(X, Y, num.cores=no_cores, cat.map=cat.map, FUN=RandMatRF, store.impurity=TRUE)
feature.imp.R <- FeatureImportance(forest, num.cores=1L, type="R")
feature.imp.C <- FeatureImportance(forest, num.cores=1L, type="C")

saveRDS(list(R=feature.imp.R, C=feature.imp.C), '../data/rf/feature_imp.rds')
```

## Cross Validate by Year

```{r xv_year, eval=FALSE}
require(rerf)
require(tidyverse)
require(lolR)
require(ggplot2)
require(mltools)
require(parallel)
require(data.table)

no_cores <- detectCores() - 1  # number of cores to parallelize over
# years <- unique(X$year)
# cross validation loop
# read in the prepared cross-validation data
inobj <- readRDS('../data/rf/rf_ins.rds')

X <- inobj$X; Y <- inobj$Y; cat.map <- inobj$cat.map
years <- sort(unique(X$year))
Y.labs <- unique(Y)
res.xv <- lapply(1:length(years), function(i) {
  yr <- years[i]
  print(years[i])
  # compute the held-out year
  tr.idx <- which(X$year != yr); test.idx <- which(X$year == yr)
  X.train <- X[tr.idx,] %>% select(-year); Y.train <- Y[tr.idx];
  X.test <- X[test.idx,] %>% select(-year); Y.test <- Y[test.idx]
  # train the forest
  forest <- RerF(X.train, Y.train, num.cores=no_cores, FUN=RandMatRF, cat.map = cat.map)
  # get the predictions on the held-out data
  Y.hat <- rerf::Predict(X.test, forest, num.cores=1L, Xtrain = X.train)
  # save the forest for shiny usage
  saveRDS(forest, sprintf('../data/rf/trained_rf_year-%d.rds', yr))
  conf.mtx <- lapply(Y.labs, function(y.true) {
    # obtain indices corresponding to the true label
    Y.test.true <- which(Y.test == y.true)
    # obtain true counts
    n.true <- length(Y.test.true)
    lapply(Y.labs, function(y.pred) {
      Y.test.pred <- which(Y.hat == y.pred)
      # which indices with true label of y.true obtain a prediction of y.pred
      Y.hat.true.pred <- intersect(Y.test.true, Y.test.pred)
      # obtain count that receive a particular prediction
      n.true.pred <- length(Y.hat.true.pred)
      
      # record count and percentage
      return(data.frame(fold=yr, True=y.true, Prediction=y.pred, n=n.true, n.pred=n.true.pred,
                        percent=n.true.pred/n.true))
    }) %>%
      bind_rows()
  }) %>%
    bind_rows()
  acc.rt <- mean(Y.hat == Y.test)
  # compare accuracy against the chance classifier which just guesses the maximally present class
  # in the training data
  labs.ct <- (Y.train %>% table())
  Y.max <- names(labs.ct)[which(labs.ct == max(labs.ct))]
  dat.overall <- data.frame(fold=yr, accuracy=acc.rt, chance=mean(Y.test == Y.max))
  return(list(conf.mtx=conf.mtx, forest=forest, dat.overall=dat.overall))
})
# aggregate
conf.mtx <- lapply(res.xv, function(res) res$conf.mtx) %>%
  bind_rows()

dat.overall <- lapply(res.xv, function(res) res$dat.overall) %>%
  bind_rows()

forests <- lapply(res.xv, function(res) res$forest)
names(forests) <- paste("forest.", test, sep="")
# save
saveRDS(list(forests=conf.mtx=conf.mtx, overall=dat.overall),
        '../data/rf/rf_results_year.rds')

saveRDS(forests, '../data/rf/trained_model_by-year.rds')
```

# Evaluation

```{r eval_scripts}
folds.res <- readRDS('../data/rf/rf_results_folds.rds')

years.res <- readRDS('../data/rf/rf_results_year.rds')
# a function for generating a confusion matrix
conf.mtx.plot <- function(dat) {
  dat %>%
    # group by the true labels and the predicted labels
    group_by(True, Prediction) %>%
    # summarize with the means per true/predicted pair
    summarize(Proportion=mean(percent), Std=sd(percent)) %>%
    # construct label for each cell of confusion matrix
    mutate(Percentage=sprintf("%.1f±%.1f %%", 100*Proportion, 100*Std)) %>%
    ggplot(aes(x=True, y=Prediction, fill=100*Proportion)) +
      geom_tile() +
      geom_text(aes(x = True, y = Prediction, label=Percentage)) +
      xlab("True Class") +
      ylab("Predicted Class") +
      ggtitle("50-Fold Cross Validation Confusion Matrix") +
      scale_fill_gradient(name="Percent", low="#FFFFFF", high="#BB00FF", limits=c(0, 100)) +
      theme_bw()
}
```

## Confusion Matrices and Predictive Accuracy

### 50-Fold Cross Validation

```{r 50fold_conf}
# construct a png and print the confusion matrix to it
png("./figs/xv_folds.png", width=450, height=370)
print(conf.mtx.plot(folds.res$conf.mtx))
dev.off()
```

### Leave-One-Year-Out Validation

## Feature Importances

```{r feature_imp}
feature.imp.raw <- readRDS('../data/rf/feature_imp.rds')
# read in the X, Y so we know what feature ids correspond to
feat.names <- colnames(readRDS('../data/rf/rf_ins.rds')$X)
```

```{r feature_imp_plt}
png("./figs/feat_imp.png", width=450, height=370)
print(data.frame(Feature.idx=sapply(feature.imp.raw$R$features, function(feat) feat[1]),
                            Gini=feature.imp.raw$R$imp) %>%
  mutate(Feature=feat.names[Feature.idx]) %>%
  left_join(data.frame(Feature.idx=sapply(feature.imp.raw$C$features, function(feat) feat[1]),
              Count=feature.imp.raw$C$imp), by="Feature.idx") %>%
  mutate(Gini=Gini/max(Gini), Count=Count/max(Count)) %>%
  top_n(10, wt=Gini) %>%
  select(Feature, Gini, Count) %>%
  gather(Metric, Measure, Gini:Count) %>%
  mutate(Feature=recode_factor(Feature, `launch_angle`="Launch Angle", `launch_speed`="Launch Speed",
                               "spray_angle"="Spray Angle", "plate_x"="Plate, Xpos", "plate_z"="Plate, Zpos",
                               "release_spin_rate"="Release Spin Rate", "ay"="Acceleration, y",
                               "sz_top"="Top of Strike Zone", "release_pos_z"="Release Position, Z",
                               "sz_bot"="Bottom of Strike Zone")) %>%
  ggplot(aes(x=Feature, y=Measure, group=Metric, fill=Metric)) +
    geom_col(position="dodge") +
    theme_bw() +
    xlab("Feature Name") +
    ylab("Normalized Metric") +
    ggtitle("Feature Importances, Top 10") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)))
dev.off()
```
